DANA Normalization Assessment for TCGA-UCEC Data

### 2021-12-17 TCGA_UCEC_Data_Analysis.Rmd
Major Revision
Considerably improved the DANA method by replacing the metric for removal of handling effects by the mean-squared correlation reduction in negative controls. Updated figures and results

---
title: "Assessment of Normalization Methods for TCGA-UCEC Data"
author: "Yannick Dueren, Johannes Lederer, Li-Xuan Qin"
output: 
  html_document:
    toc: true
    toc_depth: 4
    toc_float: true
# abstract: "TODO: This is the abstract."
knit: (function(inputFile, encoding) {
  rmarkdown::render(inputFile, encoding = encoding, output_dir = "results") })
---

@@

The settings list `settings` configure which parts, simulations, and analysis steps of this R markdown are executed. This way single components of the analysis can be disabled for the purpose of saving computation time. Setting `settings$all` to `TRUE` will set all other options to true regardless of their value.

```{r analysis.settings}
settings <- list(
  # If true sets all other settings to TRUE
  all=F,
  # Application of the full DANA pipeline to the benchmark data
  bench.DANA=T,
  # Application of the full DANA pipeline to the test data
  test.DANA=T,
  # Generate interactive plots of co-expression structures
  investigate.coexpression=T,
  # Perform Differential Expression Analysis for test and benchmark data sets
  perform.DEA=T,
  # Generate paper figures
  generate.paper.figs=T,
  # Specify if paper figures are exported
  export.figures = T,
  # Path for file exports
  paper.fig.path = "../danawriteup/figs/",
  # Clears environment
  debug=TRUE
)
if(settings$all) {
  settings$bench.DANA=TRUE
  settings$test.DANA=TRUE
  settings$investigate.coexpression=TRUE
  settings$generate.paper.figs=TRUE
}
if(settings$generate.paper.figs) {
  settings$bench.DANA=TRUE
  settings$test.DANA=TRUE
}
if(settings$debug) rm(list=ls()[ls()!="settings"])
```

@@

  # Combined read count histogram for test and benchmark data
  df <- data.frame(bench=bench.RC.log2, test=test.RC.log2)
  p.RC.hist <- ggplot(df) + 
    theme_classic() +
    geom_histogram(aes(x=test, y=..count..,fill="#69b3a2"), binwidth = .1) +
    geom_histogram(aes(x=bench, y=-..count.., fill="#404080"), binwidth = .1) +
    geom_vline(xintercept = log2(config$t.zero+1), color="#5851b8", linetype="twodash") +
    geom_vline(xintercept = log2(config$t.poor+1), color="#5851b8", linetype="twodash") +
    geom_vline(xintercept = log2(config$t.well+1), color="#E7298A",  linetype="longdash") +
    geom_segment(aes(x = log2(config$t.zero+1), y = 180, xend = log2(config$t.poor+1), yend = 180),
                 arrow=arrow(length=unit(.07, "in"), ends="both"),
                 color="#5851b8") +
    annotate(geom="label", x=(log2(config$t.zero+1)+log2(config$t.poor+1))/2.0, y=200,
             label="neg. contr.", color="#5851b8", size=4, fill = alpha(c("white"),0.85), label.size = NA) +
    geom_segment(aes(x = log2(config$t.well+1), y = 180, xend = 10, yend = 180),
                 arrow=arrow(length=unit(.07, "in"), ends="last"),
                 color="#E7298A") +
    annotate(geom="label", x=log2(config$t.well+1)+.5, y=200,
             label="pos. contr.", color="#E7298A", size=4, hjust=0, fill = alpha(c("white"),0.85), label.size = NA) +
    # ylim(c(-90,90)) +
    scale_fill_manual(name="Data set",values=c("#404080", "#69b3a2"), labels=c("single-batch", "mixed-batch"), guide = guide_legend(reverse=TRUE)) +
    # ggtitle("log2(Read count + 1)") +
    theme(legend.position=c(0.86,0.86), 
          # legend.title=element_blank(),
          legend.background=element_rect(fill=alpha('white', 0.9))) +
    scale_y_continuous(breaks=c(-200,-150,-100,-50,0,50,100,150,200), labels=c(200,150,100,50,0,50,100,150,200), limits=c(-200,200)) +
    xlab("Mean log2(read count)") +
    ylab("Frequency")
  print(p.RC.hist)
  rm(df)

@@

  # Common control miRNAs
  pos.controls.bench <- rownames(bench.model$pos$corr)
  pos.controls.test <- rownames(test.model$pos$corr)
  pos.controls.common <- intersect(pos.controls.test, pos.controls.bench)
  neg.controls.bench <- rownames(bench.model$neg$corr)
  neg.controls.test <- rownames(test.model$neg$corr)
  # neg.controls.common <- intersect(neg.controls.test, neg.controls.bench)
  # reduce the number of genes for the plot
  num.genes.plot.pos <- min(num.genes.plot.pos, length(pos.controls.common))
  pos.controls.plot <- pos.controls.common[1:num.genes.plot.pos]

@@

  p.corr.pos <- arrangeGrob(p.corr.pos.bench + theme(legend.position="none"), 
                            p.corr.pos.test + theme(legend.position="none"),
                            get.legend(p.corr.pos.bench), 
                            widths = c(2,2,1))
  grid.arrange(p.corr.pos)  # plot

  p.corr.pos.three <- arrangeGrob(p.corr.pos.bench + theme(legend.position="none"), 
                                  p.corr.pos.test.TMM + theme(legend.position="none"),
                                  p.corr.pos.test + theme(legend.position="none"),
                                  get.legend(p.corr.pos.bench + theme(legend.position = "bottom")),
                                  layout_matrix=rbind(c(1,2,3), c(4,4,4)),
                                  heights = c(5,1))
  grid.arrange(p.corr.pos.three)  # plot

@@

  p.corr.neg <- arrangeGrob(p.corr.neg.bench + theme(legend.position="none"), 
                            p.corr.neg.test + theme(legend.position="none"),
                            get.legend(p.corr.neg.bench),
                            widths = c(2,2,1))
  grid.arrange(p.corr.neg)  # plot

  p.corr.neg.three <- arrangeGrob(p.corr.neg.bench + theme(legend.position="none"), 
                                  p.corr.neg.test.TMM + theme(legend.position="none"),
                                  p.corr.neg.test + theme(legend.position="none"),
                                  get.legend(p.corr.neg.bench + theme(legend.position = "bottom")),
                                  layout_matrix=rbind(c(1,1,2,2,3,3), c(4,4,4,4,4,4)),
                                  heights = c(5,1))
  grid.arrange(p.corr.neg.three)  # plot

@@

### Subset for paper

```{r generate.paper.figs.corr.density}
if(settings$generate.paper.figs) {
  # Co-expression models
  bench.model <- DANA.bench$data.model
  bench.norm.model <- DANA.bench$norm.models$benchmark.TMM
  test.model <- DANA.test$data.model
  test.norm.model1 <- DANA.test$norm.models$test.TMM
  test.norm.model2 <- DANA.test$norm.models$test.TC

  # Set number of genes for negative correlation to minimum
  num.genes.plot.neg <- min(dim(test.model$neg$corr)[1],
                            dim(bench.model$neg$corr)[1],
                            dim(test.norm.model1$neg$corr)[1],
                            dim(test.norm.model2$neg$corr)[1])

  # Subsetted correlation matrices
  corr.bench.neg <- bench.model$neg$corr[1:num.genes.plot.neg, 1:num.genes.plot.neg]
  corr.bench.norm1.neg <- bench.norm.model$neg$corr[1:num.genes.plot.neg, 1:num.genes.plot.neg]
  corr.test.neg <- test.model$neg$corr[1:num.genes.plot.neg, 1:num.genes.plot.neg]
  corr.test.norm1.neg <- test.norm.model1$neg$corr[1:num.genes.plot.neg, 1:num.genes.plot.neg]
  corr.test.norm2.neg <- test.norm.model2$neg$corr[1:num.genes.plot.neg, 1:num.genes.plot.neg]

  ## extract non-zero correlations
  nnz.corr.test.neg <- corr.test.neg[upper.tri(corr.test.neg)]
  nnz.corr.test.neg <- nnz.corr.test.neg[nnz.corr.test.neg != 0]
  nnz.corr.test.norm1.neg <- corr.test.norm1.neg[upper.tri(corr.test.norm1.neg)]
  nnz.corr.test.norm1.neg <- nnz.corr.test.norm1.neg[nnz.corr.test.norm1.neg != 0]
  nnz.corr.test.norm2.neg <- corr.test.norm2.neg[upper.tri(corr.test.norm2.neg)]
  nnz.corr.test.norm2.neg <- nnz.corr.test.norm2.neg[nnz.corr.test.norm2.neg != 0]
  nnz.corr.bench.neg <- corr.bench.neg[upper.tri(corr.bench.neg)]
  nnz.corr.bench.neg <- nnz.corr.bench.neg[nnz.corr.bench.neg != 0]
  nnz.corr.bench.norm1.neg <- corr.bench.norm1.neg[upper.tri(corr.bench.norm1.neg)]
  nnz.corr.bench.norm1.neg <- nnz.corr.bench.norm1.neg[nnz.corr.bench.norm1.neg != 0]

  # Keep positive correlations
  nnz.corr.test.neg <- nnz.corr.test.neg[nnz.corr.test.neg > 0]
  nnz.corr.test.norm1.neg <- nnz.corr.test.norm1.neg[nnz.corr.test.norm1.neg > 0]
  nnz.corr.test.norm2.neg <- nnz.corr.test.norm2.neg[nnz.corr.test.norm2.neg > 0]
  nnz.corr.bench.neg <- nnz.corr.bench.neg[nnz.corr.bench.neg > 0]
  nnz.corr.bench.norm1.neg <- nnz.corr.bench.norm1.neg[nnz.corr.bench.norm1.neg > 0]

  ## direct comparison of negative controls benchmark vs test vs test.norm1
  neg.corr <- data.frame(
    control=factor(c(rep("Single-batch", length(nnz.corr.bench.neg)), 
                     rep("Mixed-batch", length(nnz.corr.test.neg)),
                     rep("Mixed-batch (TMM)", length(nnz.corr.test.norm1.neg)))),
    corr=abs(c(nnz.corr.bench.neg, nnz.corr.test.neg, nnz.corr.test.norm1.neg))
    )

  p.density.neg.nnz.freq <- ggplot(neg.corr, aes(x=corr, colour=control, linetype=control)) +
    geom_freqpoly(binwidth=.05, alpha=.8, size=.9) +
    theme_classic() +
    xlim(0, 1) +
    scale_linetype_manual(values=c("solid", "solid", "twodash"))+
    scale_color_manual(values=c('#1B9E77','#D95F02','#D95F02')) + # First 2 colors from rcolorbrewer set "Dark2"
    theme(legend.position=c(0.9,0.9), 
          legend.title=element_blank(),
          legend.background=element_rect(fill=alpha('white', 0.5))) +
    ylab("Frequency") +
    xlab("Marginal correlation") +
    ggtitle("Positive correlation among negative controls")
  print(p.density.neg.nnz.freq)

  ## direct comparison of negative controls Single-batch vs test vs test.norm1 vs test.norm2
  neg.corr <- data.frame(
    control=factor(c(rep("Single-batch", length(nnz.corr.bench.neg)), 
                     rep("Mixed-batch", length(nnz.corr.test.neg)),
                     rep("Mixed-batch (TMM)", length(nnz.corr.test.norm1.neg)),
                     rep("Mixed-batch (TC)", length(nnz.corr.test.norm2.neg)))),
    corr=abs(c(nnz.corr.bench.neg, nnz.corr.test.neg, nnz.corr.test.norm1.neg, nnz.corr.test.norm2.neg))
    )
  # ttt <- data.frame(prop.table(table(method=neg.corr$control, corr=neg.corr$corr), 1))

  p.density.neg.nnz.freq <- ggplot(neg.corr, aes(x=corr, colour=control, linetype=control)) +
    geom_freqpoly(binwidth=.05, alpha=.8, size=.9) +
    theme_classic() +
    xlim(0, 1) +
    # scale_linetype_manual(values=c("twodash", "solid", "solid", "solid"))+
    scale_color_manual(values=c('#e7298a','#1B9E77','#D95F02', '#7570B3')) + # First colors from rcolorbrewer set "Dark2"
    theme(legend.position=c(0.9,0.86), 
          legend.title=element_blank(),
          legend.background=element_rect(fill=alpha('white', 0.5))) +
    ylab("Frequency") +
    xlab("Marginal correlation")#+ ggtitle("Positive correlation among negative controls")
  print(p.density.neg.nnz.freq)

  # Additional Plot with normalized Single-batch
  neg.corr <- data.frame(
    control=factor(c(rep("Single-batch", length(nnz.corr.bench.neg)), 
                     rep("Single-batch (TMM)", length(nnz.corr.bench.norm1.neg)),
                     rep("Mixed-batch", length(nnz.corr.test.neg)),
                     rep("Mixed-batch (TMM)", length(nnz.corr.test.norm1.neg)))),
    corr=abs(c(nnz.corr.bench.neg, nnz.corr.bench.norm1.neg, nnz.corr.test.neg, nnz.corr.test.norm1.neg))
    )
  p.density.neg.nnz.freq.four <- ggplot(neg.corr, aes(x=corr, colour=control, linetype=control)) +
    geom_freqpoly(binwidth=.05, alpha=.8, size=1) +
    theme_classic() +
    xlim(0, 1) +
    theme(legend.position=c(0.9,0.9), 
          legend.title=element_blank(),
          legend.background=element_rect(fill=alpha('white', 0.5))) +
    ylab("Frequency") +
    xlab("Marginal correlation") +
    scale_linetype_manual(values=c("solid", "twodash","solid", "twodash"))+
    scale_color_manual(values=c('#1B9E77','#1B9E77','#D95F02','#D95F02')) + # First 2 colors from rcolorbrewer set "Dark2"
    ggtitle("Positive correlation among negative controls")
  print(p.density.neg.nnz.freq.four)

  if(settings$export.figures) {
    ggsave(paste0(settings$paper.fig.path, "UCEC_CorrDensityNeg_Freq.pdf"), p.density.neg.nnz.freq, width = 5, height=4, device="pdf")
    ggsave(paste0(settings$paper.fig.path, "UCEC_CorrDensityNeg_Freq_WithBench.pdf"), p.density.neg.nnz.freq.four, width = 5, height=4, device="pdf")
  }
}
```

### All methods

```{r generate.paper.figs.corr.density.supplements}
if(settings$generate.paper.figs) {
  # Co-expression models
  bench.model <- DANA.bench$data.model
  test.model <- DANA.test$data.model
  test.norm.model1 <- DANA.test$norm.models$test.TMM
  test.norm.model2 <- DANA.test$norm.models$test.RUVs

  # Set number of genes for negative correlation to minimum
  num.genes.plot.neg <- min(dim(test.model$neg$corr)[1],
                            dim(bench.model$neg$corr)[1],
                            sapply(DANA.test$norm.models, function(x) dim(x$neg$corr)[1]))

  # Subsetted correlation matrices
  corr.bench.neg <- bench.model$neg$corr[1:num.genes.plot.neg, 1:num.genes.plot.neg]
  corr.test.neg <- test.model$neg$corr[1:num.genes.plot.neg, 1:num.genes.plot.neg]
  corr.test.norm.neg <- lapply(DANA.test$norm.models, function(x) x$neg$corr[1:num.genes.plot.neg, 1:num.genes.plot.neg])

  ## Strictly positive correlations
  nnz.corr.bench.neg <- corr.bench.neg[upper.tri(corr.bench.neg)]
  nnz.corr.bench.neg <- nnz.corr.bench.neg[nnz.corr.bench.neg > 0]
  nnz.corr.bench.norm1.neg <- corr.bench.norm1.neg[upper.tri(corr.bench.norm1.neg)]
  nnz.corr.bench.norm1.neg <- nnz.corr.bench.norm1.neg[nnz.corr.bench.norm1.neg > 0]
  nnz.corr.test.neg <- corr.test.neg[upper.tri(corr.test.neg)]
  nnz.corr.test.neg <- nnz.corr.test.neg[nnz.corr.test.neg > 0]
  nnz.corr.test.norm.neg <- lapply(corr.test.norm.neg, function(x) x[upper.tri(x)][x[upper.tri(x)] > 0])
  names(nnz.corr.test.norm.neg) <- substr(names(nnz.corr.test.norm.neg), 6, 20)

  control <- c(rep("Single-batch", length(nnz.corr.bench.neg)), 
                     rep("Mixed-batch", length(nnz.corr.test.neg)))
  for(i in 1:length(nnz.corr.test.norm.neg)) {
    control <- c(control, rep(paste0("Mixed-batch (", names(nnz.corr.test.norm.neg)[i],")"), length(nnz.corr.test.norm.neg[[i]])))
  }
  ## direct comparison of negative controls Single-batch vs test vs test.norm1
  neg.corr <- data.frame(
    control=factor(control),
    corr=abs(c(nnz.corr.bench.neg, nnz.corr.test.neg, unname(unlist(nnz.corr.test.norm.neg))))
    )

  p.corr.frequency.neg.controls.all <- ggplot(neg.corr, aes(x=corr, colour=control, linetype=control)) +
    geom_freqpoly(binwidth=.05, alpha=.9, size=.9) +
    theme_classic() +
    xlim(0, 1) +
    scale_color_manual(values=brewer.pal(n=12, name="Paired")) + # First 2 colors from rcolorbrewer set "Dark2"
    theme(legend.position=c(0.85,0.6), 
          legend.title=element_blank(),
          legend.background=element_rect(fill=alpha('white', 0.5))) +
    ylab("Frequency") +
    xlab("Marginal correlation") + # ggtitle("Positive correlation among negative controls") + 
    theme(legend.key.width = unit(2,"cm"))
  print(p.corr.frequency.neg.controls.all)

  if(settings$export.figures) {
    ggsave(paste0(settings$paper.fig.path, "UCEC_CorrDensityNeg_Freq_All.pdf"), p.corr.frequency.neg.controls.all, width = 8, height=5, device="pdf")
  }
}
```

@@

### Subset for paper

```{r generate.paper.figs.corr.scatter}
if(settings$generate.paper.figs) {
  ## Generate correlation scatter plots for positive controls for TMM and UQ
  par(mfrow=c(1,1))

  # un-normalized vs TMM
  p.scatter.test.TMM <- plot.corr.scatter(
    corr1=DANA.test$data.model$pos$corr, 
    corr2=DANA.test$norm.models$test.TMM$pos$corr,
    clusters=clusters.test,
    title="Mixed-batch (TMM)",
    xlab="un-normalized", 
    ylab="TMM", 
    threshold=0)
  print(p.scatter.test.TMM)
  # un-normalized vs RUVs
  p.scatter.test.RUVs <- plot.corr.scatter(
    corr1=DANA.test$data.model$pos$corr, 
    corr2=DANA.test$norm.models$test.RUVs$pos$corr,
    clusters=clusters.test,
    title="Mixed-batch (RUVs)",
    xlab="un-normalized", 
    ylab="RUVs", 
    threshold=0)
  print(p.scatter.test.RUVs)

  # Save plots
  if(settings$export.figures) {
    ggsave(paste0(settings$paper.fig.path, "UCEC_ScatterTMM.pdf"), p.scatter.test.TMM, width=4, height=4, device="pdf")
    ggsave(paste0(settings$paper.fig.path, "UCEC_ScatterRUVs.pdf"), p.scatter.test.RUVs, width=4, height=4, device="pdf")
    # ggsave(paste0(settings$paper.fig.path, "UCEC_CorrNeg.eps"), p.corr.neg, width = 8, height=3.5, device="eps")
  }
}
```

### All normalization methods for supplementary materials

@@

  test.DANA.metrics <- data.frame(
    method=substr(rownames(DANA.test$metrics), 6, 20),
    cc=DANA.test$metrics[, "cc"],
    mcr=DANA.test$metrics[, "mcr"]
  )

  p.metrics.test <- ggplot(test.DANA.metrics,aes(x=mcr,y=cc, label=method)) + 
    geom_point(alpha=.5) + #pos=position_jitter(width=0.001, seed=1),
    theme_classic() + 
    xlab(TeX("$mcr^-$; Relative reduction of handling effects")) + 
    ylab(TeX("$cc^+$; Biological signal preservation")) + 
    geom_text_repel(aes(label = method), size=3, max.overlaps = Inf, min.segment.length = 0.2, force = 1, box.padding = 0.2) +
    # xlim(c(min(c(0, test.mposc.reduction)),max(test.mposc.reduction))) +
    scale_x_continuous(labels = scales::percent, limits=c(0,1.05), 
                       breaks = scales::pretty_breaks(n = 5)) +
    # ylim(c(0,1)) 
    scale_y_continuous(labels = scales::percent, limits = c(0,1))
  print(p.metrics.test)

  print("DANA Statistics for mixed-batch data")
  test.DANA.metrics$cc <- round(test.DANA.metrics$cc,3)
  test.DANA.metrics$mcr <- round(test.DANA.metrics$mcr,3)
  print(test.DANA.metrics)

  bench.DANA.metrics <- data.frame(
    method=substr(rownames(DANA.bench$metrics), 11, 30),
    cc=DANA.bench$metrics[, "cc"],
    mcr=DANA.bench$metrics[, "mcr"]
  )

  p.metrics.bench <- ggplot(bench.DANA.metrics, aes(x=mcr, y=cc, label=method)) + 
    geom_point(alpha=.5) +
    theme_classic() + 
    xlab(TeX("$mcr^-$; Relative reduction of handling effects")) + 
    ylab(TeX("$cc^+$; Biological signal preservation")) + 
    geom_text_repel(aes(label = method), size=3) +
    # xlim(c(min(c(0,bench.mposc.reduction)),max(bench.mposc.reduction))) +
    scale_x_continuous(labels = scales::percent, limits=c(0, 1.05),
                       breaks = scales::pretty_breaks(n = 5)) +
    # ylim(c(0,1))
    scale_y_continuous(labels = scales::percent, limits = c(0, 1)) 
  print(p.metrics.bench)

  print("DANA Statistics for single-batch data")
  bench.DANA.metrics$cc <- round(bench.DANA.metrics$cc,3)
  bench.DANA.metrics$mcr <- round(bench.DANA.metrics$mcr,3)
  print(bench.DANA.metrics)

  if(settings$export.figures) {
    ggsave(paste0(settings$paper.fig.path, "UCEC_MetricsTest.pdf"), p.metrics.test, width=4.5, height=3.5, device="pdf")
    ggsave(paste0(settings$paper.fig.path, "UCEC_MetricsBench.pdf"), p.metrics.bench, width=4.5, height=3.5, device="pdf")
  }

@@

```{r generate.paper.figs.DEA}
if(settings$generate.paper.figs && settings$perform.DEA) { 
  # FNR-FDR Plot
  test.DEA.res <- data.frame(method=substr(names(test.norm.DEA.stats), 6, 20),
                             FDR=-sapply(test.norm.DEA.stats, function(x) x$FDR)+1,
                             FNR=-sapply(test.norm.DEA.stats, function(x) x$FNR)+1)
  p.test.DEA.FNR.FDR <- ggplot(test.DEA.res, aes(x=FNR, y=FDR, label=method)) +
      theme_classic() + 
      geom_point(alpha=.5) + 
      xlab("Sensitivity") + # True positive rate(1-FNR)
      ylab("Precision") + # Positive predictive value (1-FDR)
      geom_text_repel(aes(label = method), size=3, max.overlaps = Inf) + 
      scale_x_continuous(labels = scales::percent, limits=c(0,1),
                         breaks = scales::pretty_breaks(n = 4)) +
      scale_y_continuous(labels = scales::percent, limits = c(0,1))
  print(p.test.DEA.FNR.FDR)

>>>>

---
title: "DANA Normalization Assessment for TCGA-UCEC Data"
author: "Yannick Dueren, Johannes Lederer, Li-Xuan Qin"
output: 
  html_document:
    toc: true
    toc_depth: 4
    toc_float: true
knit: (function(inputFile, encoding) {
  rmarkdown::render(inputFile, encoding = encoding, output_dir = "results") })
---

@@

The settings list `settings` configure which parts, simulations, and analysis steps of this R markdown are executed. This way single components of the analysis can be disabled for the purpose of saving computation time.

```{r analysis.settings}
settings <- list(
  # Application of the full DANA pipeline to the benchmark data
  bench.DANA=T,
  # Application of the full DANA pipeline to the test data
  test.DANA=T,
  # Generate interactive plots of co-expression structures
  investigate.coexpression=F,
  # Perform Differential Expression Analysis for test and benchmark data sets
  perform.DEA=T,
  # Generate paper figures
  generate.paper.figs=T,
  # Specify if paper figures are exported
  export.figures = T,
  # Path for file exports
  paper.fig.path = "../danawriteup/figs/",
  # Clears environment
  debug=TRUE
)
if(settings$debug) rm(list=ls()[ls()!="settings"])
```

@@

  # Combined read count histogram for test and benchmark data
  df <- data.frame(bench=bench.RC.log2, test=test.RC.log2)
  p.RC.hist <- ggplot(df) + 
    theme_classic() +
    geom_histogram(aes(x=test, y=..count..,fill="#69b3a2"), binwidth = .1) +
    geom_histogram(aes(x=bench, y=-..count.., fill="#404080"), binwidth = .1) +
    geom_vline(xintercept = log2(config$t.zero+1), color="#5851b8", linetype="twodash") +
    geom_vline(xintercept = log2(config$t.poor+1), color="#5851b8", linetype="twodash") +
    geom_vline(xintercept = log2(config$t.well+1), color="#E7298A",  linetype="longdash") +
    geom_segment(aes(x = log2(config$t.zero+1), y = 180, xend = log2(config$t.poor+1), yend = 180),
                 arrow=arrow(length=unit(.07, "in"), ends="both"),
                 color="#5851b8") +
    annotate(geom="label", x=(log2(config$t.zero+1)+log2(config$t.poor+1))/2.0, y=200,
             label="neg. contr.", color="#5851b8", size=4, fill = alpha(c("white"),0.85), label.size = NA) +
    geom_segment(aes(x = log2(config$t.well+1), y = 180, xend = 10, yend = 180),
                 arrow=arrow(length=unit(.07, "in"), ends="last"),
                 color="#E7298A") +
    annotate(geom="label", x=log2(config$t.well+1)+.5, y=200,
             label="pos. contr.", color="#E7298A", size=4, hjust=0, fill = alpha(c("white"),0.85), label.size = NA) +
    scale_fill_manual(name="Data set",values=c("#404080", "#69b3a2"), labels=c("single-batch", "mixed-batch"), guide = guide_legend(reverse=TRUE)) +
    theme(legend.position=c(0.86,0.86), 
          legend.background=element_rect(fill=alpha('white', 0.9))) +
    scale_y_continuous(breaks=c(-200,-150,-100,-50,0,50,100,150,200), labels=c(200,150,100,50,0,50,100,150,200), limits=c(-200,200)) +
    xlab("Mean log2(read count)") +
    ylab("Frequency")
  print(p.RC.hist)
  rm(df)

@@

  # Common control miRNAs
  pos.controls.bench <- rownames(bench.model$pos$corr)
  pos.controls.test <- rownames(test.model$pos$corr)
  pos.controls.common <- intersect(pos.controls.test, pos.controls.bench)
  neg.controls.bench <- rownames(bench.model$neg$corr)
  neg.controls.test <- rownames(test.model$neg$corr)
  # reduce the number of genes for the plot
  num.genes.plot.pos <- min(num.genes.plot.pos, length(pos.controls.common))
  pos.controls.plot <- pos.controls.common[1:num.genes.plot.pos]

@@

  p.corr.pos <- arrangeGrob(p.corr.pos.bench + theme(legend.position="none"), 
                            p.corr.pos.test + theme(legend.position="none"),
                            get.legend(p.corr.pos.bench), 
                            widths = c(2,2,1))
  grid.arrange(p.corr.pos)  # display plot

  p.corr.pos.three <- arrangeGrob(p.corr.pos.bench + theme(legend.position="none"), 
                                  p.corr.pos.test.TMM + theme(legend.position="none"),
                                  p.corr.pos.test + theme(legend.position="none"),
                                  get.legend(p.corr.pos.bench + theme(legend.position = "bottom")),
                                  layout_matrix=rbind(c(1,2,3), c(4,4,4)),
                                  heights = c(5,1))
  grid.arrange(p.corr.pos.three)  # display plot

@@

  p.corr.neg <- arrangeGrob(p.corr.neg.bench + theme(legend.position="none"), 
                            p.corr.neg.test + theme(legend.position="none"),
                            get.legend(p.corr.neg.bench),
                            widths = c(2,2,1))
  grid.arrange(p.corr.neg)  # display plot

  p.corr.neg.three <- arrangeGrob(p.corr.neg.bench + theme(legend.position="none"), 
                                  p.corr.neg.test.TMM + theme(legend.position="none"),
                                  p.corr.neg.test + theme(legend.position="none"),
                                  get.legend(p.corr.neg.bench + theme(legend.position = "bottom")),
                                  layout_matrix=rbind(c(1,1,2,2,3,3), c(4,4,4,4,4,4)),
                                  heights = c(5,1))
  grid.arrange(p.corr.neg.three)  # display plot

@@

```{r generate.paper.figs.corr.density.supplements}
if(settings$generate.paper.figs) {
  models <- list(
    "Single-batch" = DANA.bench$data.model,
    "Mixed-batch" = DANA.test$data.model,
    "Mixed-batch (TMM)" = DANA.test$norm.models$test.TMM,
    "Mixed-batch (DESeq)" = DANA.test$norm.models$test.DESeq,
    "Mixed-batch (PoissonSeq)" = DANA.test$norm.models$test.PoissonSeq,
    "Mixed-batch (RUVg)" = DANA.test$norm.models$test.RUVg,
    "Mixed-batch (RUVs)" = DANA.test$norm.models$test.RUVs,
    "Mixed-batch (RUVr)" = DANA.test$norm.models$test.RUVr,
    "Mixed-batch (TC)" = DANA.test$norm.models$test.TC,
    "Mixed-batch (Med)" = DANA.test$norm.models$test.Med,
    "Mixed-batch (UQ)" = DANA.test$norm.models$test.UQ,
    "Mixed-batch (QN)" = DANA.test$norm.models$test.QN
  )

  # Set number of genes for negative correlation to minimum found
  num.genes.plot <- min(sapply(models, function(x) dim(x$neg$corr)[1]))

  # Subsetted correlation matrices
  corrs <- lapply(models, function(x) x$neg$corr[1:num.genes.plot, 1:num.genes.plot])
  corrs <- lapply(corrs, function(x) x[upper.tri(x)])

  control <- factor(
    as.vector(sapply(names(corrs), function(x) rep(x,length(corrs[[1]])))),
    levels = names(models) 
  )
  neg.corr <- data.frame(
    control=factor(control),
    corr=unname(unlist(corrs))
  )

  # plot insert
  p.insert <- ggplot(neg.corr, aes(x=corr, colour=control, linetype=control)) +
    geom_vline(xintercept = 0, size=.4, color="gray80", linetype="solid") +
    geom_freqpoly(binwidth=.05, alpha=.9, size=.3) +
    theme_bw() +
    xlim(-1, 1) +
    scale_color_manual(values=brewer.pal(n=12, name="Paired")) +
    theme(legend.position="none",
          axis.title.x = element_blank(),
          axis.title.y = element_blank())

  p.corr.frequency.neg.controls.all <- ggplot(neg.corr, aes(x=corr, colour=control, linetype=control)) +
    geom_vline(xintercept = 0, size=.4, color="gray80", linetype="solid") +
    geom_freqpoly(binwidth=.05, alpha=.9, size=.75) +
    theme_classic() +
    xlim(-1, 1) +
    ylim(0,200) +
    scale_color_manual(values=brewer.pal(n=12, name="Paired")) +
    theme(legend.position=c(0.84,0.8), 
          legend.title=element_blank(),
          legend.background=element_rect(fill=alpha('white', 0.5))) +
    ylab("Frequency") +
    xlab("Marginal correlation") + 
    theme(legend.key.width = unit(2.4,"cm")) +
    annotation_custom(ggplotGrob(p.insert), xmin=-1.1, xmax=-0.3, ymin=100, ymax=200)
  print(p.corr.frequency.neg.controls.all)

  if(settings$export.figures) {
    ggsave(paste0(settings$paper.fig.path, "UCEC_CorrDensityNeg_Freq_All.pdf"), p.corr.frequency.neg.controls.all, width = 8, height=8, device="pdf")
  }
}
```

@@

@@

  test.DANA.metrics <- data.frame(
    method=substr(rownames(DANA.test$metrics), 6, 20),
    cc=DANA.test$metrics[, "cc"],
    mscr=DANA.test$metrics[, "mscr"]
  )

  p.metrics.test <- ggplot(test.DANA.metrics,aes(x=mscr,y=cc, label=method)) + 
    geom_point(alpha=.5) + #pos=position_jitter(width=0.001, seed=1),
    theme_classic() + 
    xlab(TeX("$mscr^-$; Relative reduction of handling effects")) + 
    ylab(TeX("$cc^+$; Biological signal preservation")) + 
    geom_text_repel(aes(label = method), size=3, max.overlaps = Inf, min.segment.length = 0.2, force = 1, box.padding = 0.2) +
    # xlim(c(min(c(0, test.mposc.reduction)),max(test.mposc.reduction))) +
    scale_x_continuous(labels = scales::percent, limits=c(0,1.05), 
                       breaks = scales::pretty_breaks(n = 5)) +
    # ylim(c(0,1)) 
    scale_y_continuous(labels = scales::percent, limits = c(0.5,1))
  print(p.metrics.test)

  print("DANA Statistics for mixed-batch data")
  test.DANA.metrics$cc <- round(test.DANA.metrics$cc,3)
  test.DANA.metrics$mscr <- round(test.DANA.metrics$mscr,3)
  print(test.DANA.metrics)

  bench.DANA.metrics <- data.frame(
    method=substr(rownames(DANA.bench$metrics), 11, 30),
    cc=DANA.bench$metrics[, "cc"],
    mscr=DANA.bench$metrics[, "mscr"]
  )

  p.metrics.bench <- ggplot(bench.DANA.metrics, aes(x=mscr, y=cc, label=method)) + 
    geom_point(alpha=.5) +
    theme_classic() + 
    xlab(TeX("$mscr^-$; Relative reduction of handling effects")) + 
    ylab(TeX("$cc^+$; Biological signal preservation")) + 
    geom_text_repel(aes(label = method), size=3) +
    scale_x_continuous(labels = scales::percent, limits=c(0, 1.05),
                       breaks = scales::pretty_breaks(n = 5)) +
    scale_y_continuous(labels = scales::percent, limits = c(0.5, 1)) 
  print(p.metrics.bench)

  print("DANA Statistics for single-batch data")
  bench.DANA.metrics$cc <- round(bench.DANA.metrics$cc,3)
  bench.DANA.metrics$mscr <- round(bench.DANA.metrics$mscr,3)
  print(bench.DANA.metrics)

  if(settings$export.figures) {
    ggsave(paste0(settings$paper.fig.path, "UCEC_MetricsTest.pdf"), p.metrics.test, width=5.5, height=4.5, device="pdf")
    ggsave(paste0(settings$paper.fig.path, "UCEC_MetricsBench.pdf"), p.metrics.bench, width=4.5, height=3.5, device="pdf")
  }

@@

```{r generate.paper.figs.DEA}
if(settings$generate.paper.figs && settings$perform.DEA) { 
  # FNR-FDR Plot
  test.DEA.res <- data.frame(method=substr(names(test.norm.DEA.stats), 6, 20),
                             FDR=-sapply(test.norm.DEA.stats, function(x) x$FDR)+1,
                             FNR=-sapply(test.norm.DEA.stats, function(x) x$FNR)+1)
  p.test.DEA.FNR.FDR <- ggplot(test.DEA.res, aes(x=FNR, y=FDR, label=method)) +
      theme_classic() + 
      geom_point(alpha=.5) + 
      xlab("Sensitivity") + # True positive rate(1-FNR)
      ylab("Positive predictive value") + # Positive predictive value (1-FDR)
      geom_text_repel(aes(label = method), size=3, max.overlaps = Inf) + 
      scale_x_continuous(labels = scales::percent, limits=c(0,1),
                         breaks = scales::pretty_breaks(n = 4)) +
      scale_y_continuous(labels = scales::percent, limits = c(0,1))
  print(p.test.DEA.FNR.FDR)
